\documentclass[11pt,a4paper]{article}

% Encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page layout
\usepackage[top=2.3cm, bottom=2.3cm, left=2.8cm, right=2.5cm]{geometry}
\setlength{\headheight}{14pt}

% Hyperlinks
\usepackage[colorlinks=true, linkcolor=black, urlcolor=blue, citecolor=black]{hyperref}

% Code listings
\usepackage{listings}
\usepackage{xcolor}

% Tables
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{tabularx}

% Typography
\usepackage{setspace}
\usepackage{parskip}
\usepackage[protrusion=true,expansion=false]{microtype}

% Headers / footers
\usepackage{fancyhdr}

% Section formatting
\usepackage{titlesec}

% ------------------------------------------------------------------ %
%  Colors
% ------------------------------------------------------------------ %
\definecolor{codegreen}{rgb}{0,0.50,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.48,0,0.62}
\definecolor{codered}{rgb}{0.58,0,0}
\definecolor{backcolour}{rgb}{0.97,0.97,0.97}

% ------------------------------------------------------------------ %
%  Listing style (numbered, for Python/proto)
% ------------------------------------------------------------------ %
\lstdefinestyle{codestyle}{
    backgroundcolor=\color{backcolour},
    commentstyle=\color{codegreen},
    keywordstyle=\color{codered}\bfseries,
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\ttfamily\footnotesize,
    breakatwhitespace=false,
    breaklines=true,
    captionpos=b,
    keepspaces=true,
    numbers=left,
    numbersep=4pt,
    showspaces=false,
    showstringspaces=false,
    tabsize=2,
    frame=single,
    framesep=3pt,
    xleftmargin=14pt,
    aboveskip=4pt,
    belowskip=4pt,
}

% Listing style (no line numbers, for protocol-level examples)
\lstdefinestyle{protstyle}{
    backgroundcolor=\color{backcolour},
    commentstyle=\color{codegreen},
    keywordstyle=\color{codered}\bfseries,
    stringstyle=\color{codepurple},
    basicstyle=\ttfamily\footnotesize,
    breakatwhitespace=false,
    breaklines=true,
    captionpos=b,
    keepspaces=true,
    numbers=none,
    showspaces=false,
    showstringspaces=false,
    tabsize=2,
    frame=single,
    framesep=3pt,
    xleftmargin=6pt,
    aboveskip=4pt,
    belowskip=4pt,
}
\lstset{style=protstyle}

% Protobuf language
\lstdefinelanguage{protobuf}{
    keywords={syntax,package,message,service,rpc,returns,stream,
              repeated,optional,required,string,int32,int64,bool,
              float,double,enum,import,option},
    keywordstyle=\color{codered}\bfseries,
    comment=[l]{//},
    morecomment=[s]{/*}{*/},
    commentstyle=\color{codegreen},
    morestring=[b]",
    stringstyle=\color{codepurple},
    sensitive=true,
}

% GraphQL language
\lstdefinelanguage{graphql}{
    keywords={type,query,mutation,subscription,input,interface,
              enum,union,scalar,fragment,on,directive,extend,schema},
    keywordstyle=\color{codered}\bfseries,
    comment=[l]{\#},
    commentstyle=\color{codegreen},
    morestring=[b]",
    stringstyle=\color{codepurple},
    sensitive=true,
}

% ------------------------------------------------------------------ %
%  Page style
% ------------------------------------------------------------------ %
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\small Comparative Analysis of API Communication Technologies}
\fancyhead[R]{\small\thepage}
\fancyfoot[C]{\small Academic Report --- February 2026}
\renewcommand{\headrulewidth}{0.4pt}
\renewcommand{\footrulewidth}{0.4pt}

% ------------------------------------------------------------------ %
%  Section title spacing (compact)
% ------------------------------------------------------------------ %
\titlespacing*{\section}{0pt}{10pt}{4pt}
\titlespacing*{\subsection}{0pt}{7pt}{3pt}
\titleformat{\section}{\large\bfseries}{\thesection}{1em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{1em}{}

% ------------------------------------------------------------------ %
%  Line / paragraph spacing
% ------------------------------------------------------------------ %
\setstretch{1.2}
\setlength{\parskip}{4pt}
\setlength{\parindent}{0pt}

% Caption spacing
\setlength{\abovecaptionskip}{3pt}
\setlength{\belowcaptionskip}{2pt}

% ================================================================== %
\begin{document}
% ================================================================== %

% ------------------------------------------------------------------ %
%  Compact title block
% ------------------------------------------------------------------ %
\begin{center}
  {\large\bfseries Comparative Analysis of Service-Oriented Communication Technologies}\\[0.3em]
  {\normalsize SOAP/WSDL \quad\textemdash\quad REST \quad\textemdash\quad
   GraphQL \quad\textemdash\quad gRPC}\\[0.25em]
  {\small Khaldoun \quad\textbar\quad
   Distributed Systems and Microservices Architecture \quad\textbar\quad
   February 2026}
\end{center}
\vspace{0.2em}
\noindent\rule{\linewidth}{0.5pt}
\vspace{-0.8em}

% ------------------------------------------------------------------ %
%  Table of contents
% ------------------------------------------------------------------ %
\setcounter{tocdepth}{2}
\tableofcontents

\vspace{0.5em}
\noindent\rule{\linewidth}{0.3pt}
\vspace{0.3em}

% ================================================================== %
\section{Introduction}
% ================================================================== %

Service-oriented computing (SOC) structures software as collections of interoperable,
loosely coupled services, each exposing a well-defined interface independently of the
underlying platform or implementation language. As distributed systems span
organisational, geographic, and technology boundaries, the choice of communication
paradigm has direct consequences for performance, security, developer experience, and
long-term maintainability.

This report compares four paradigms that have shaped distributed systems over the past
three decades. \textbf{SOAP/WSDL} emerged in the late 1990s as a standards-driven,
XML-based messaging protocol accompanied by a formal interface contract. \textbf{REST}
(Representational State Transfer) was formalised by Roy Fielding in 2000 as an
architectural style aligned with HTTP semantics, enabling simple, cacheable,
resource-oriented APIs. \textbf{GraphQL}, developed at Facebook in 2012 and open-sourced
in 2015, introduced a typed query language that allows clients to specify precisely the
data they require. \textbf{gRPC}, released by Google in 2016, uses Protocol Buffers over
HTTP/2 for high-performance binary RPC communication with native streaming support.

The analysis evaluates each technology across eight dimensions: architecture and
principles, strengths, weaknesses, technology stack and data encoding, execution
environments, security model, and use cases. Section~3 illustrates each technology with
code examples using a common user management domain. Section~4 consolidates the
comparison in a summary table, and Section~5 discusses when each approach is most
suitable.

% ================================================================== %
\section{Technology Analysis by Dimension}
% ================================================================== %

% ------------------------------------------------------------------ %
\subsection{SOAP / WSDL}
% ------------------------------------------------------------------ %

\textbf{Architecture and principles.}
SOAP (Simple Object Access Protocol) defines a strict XML-based message envelope
composed of an optional \textit{Header} --- carrying metadata such as authentication
tokens, routing directives, and transaction coordinators --- and a mandatory
\textit{Body} carrying the operation payload. The protocol is transport-agnostic but
deployed almost exclusively over HTTPS. Its companion specification, WSDL (Web Services
Description Language), provides a machine-readable XML contract that formally describes
operations, message schemas, and endpoint addresses, enabling automated client and server
stub generation. The WS-* extension family adds WS-Security (XML Signature and XML
Encryption for message-level integrity and confidentiality), WS-ReliableMessaging, and
WS-AtomicTransaction. Services are tightly coupled: WSDL changes require coordinated
updates across all consumers. Supported message exchange patterns include request-response,
one-way, solicit-response, and notification.

\textbf{Strengths.}
SOAP's defining advantage is its formalism. WS-Security provides end-to-end message-level
security independent of transport: digital signatures ensure integrity and non-repudiation;
selective encryption provides confidentiality across multi-hop intermediaries --- a
capability unique to SOAP in this comparison. The WSDL contract is machine-verifiable
and supports automated code generation via Apache CXF, JAX-WS, and WCF. WS-AtomicTransaction
enables distributed ACID transactions across organisational boundaries. These properties
make SOAP the only technology suited to contexts where message-level guarantees carry
regulatory weight.

\textbf{Weaknesses.}
XML verbosity is the primary operational cost: envelopes inflate payload size, increase
parsing latency, and complicate debugging. Despite formal standards, WS-* interoperability
between different vendor implementations has been historically inconsistent. Browser and
mobile support is minimal; debugging requires specialised XML tools rather than simple
HTTP inspection. The overall complexity of the WS-* stack raises development costs and
integration effort compared to every other option examined here.

\textbf{Stack, security, environments, and use cases.}
Server implementations include Apache CXF and JAX-WS (Java), WCF (.NET), and Spyne
(Python); messages are encoded in XML. Security is managed at message level via
WS-Security, independently of TLS. SOAP is found primarily in banking, insurance,
healthcare (HL7 v3), and government platforms where compliance, audit trails, and
transactional guarantees are legally required. It is rarely the right choice for new
projects outside these regulated contexts.

% ------------------------------------------------------------------ %
\subsection{REST (Representational State Transfer)}
% ------------------------------------------------------------------ %

\textbf{Architecture and principles.}
REST is an architectural style --- not a protocol --- defined by six constraints applied
to HTTP: client-server separation, statelessness, cacheability, uniform interface,
layered system, and optional code-on-demand. Applied to HTTP, these produce APIs where
resources are identified by URIs and manipulated via standard verbs (GET, POST, PUT,
PATCH, DELETE), with status codes providing a standardised outcome vocabulary. Statelessness
requires each request to be self-contained, enabling horizontal scaling without session
affinity. The dominant payload format is JSON, consumed by any language or platform
without specialised tooling.

\textbf{Strengths.}
Simplicity is REST's principal advantage. Any HTTP client --- a browser, cURL, or a
mobile SDK --- can interact with a REST API without code generation. GET responses are
cacheable at every HTTP layer (browser, CDN, reverse proxy), making REST exceptionally
efficient for read-heavy workloads. The OpenAPI ecosystem provides standardised
documentation, mock servers, validation, and multi-language client generation. REST's
combination of accessibility and ecosystem depth makes it the most widely adopted paradigm
for third-party API consumers.

\textbf{Weaknesses.}
REST imposes no mandatory schema, leading to inconsistent API design quality. Over-fetching
(endpoints return more data than needed) and under-fetching (multiple requests required
to assemble a view) are structural limitations that are particularly costly for mobile
clients. API versioning has no consensus strategy; URL versioning, header negotiation, and
content negotiation each carry different trade-offs. There is no standard for message-level
security.

\textbf{Stack, security, environments, and use cases.}
Frameworks include Flask and Django REST Framework (Python), Spring Boot (Java), and
Express.js (Node.js). Security relies on TLS for transport and OAuth2/JWT for
authentication. REST is the standard for public APIs, cloud provider interfaces (AWS,
GCP, Azure), SaaS platforms, and mobile backends. It is less appropriate for
high-frequency internal microservice communication, where binary protocols offer better
performance, or for clients with rapidly evolving and diverse data requirements.

% ------------------------------------------------------------------ %
\subsection{GraphQL}
% ------------------------------------------------------------------ %

\textbf{Architecture and principles.}
GraphQL exposes a single endpoint backed by a strongly typed schema defined in the
Schema Definition Language (SDL). Clients submit queries specifying precisely the fields
required; the server responds with a payload mirroring the query structure. Three root
operation types are defined: Query (read), Mutation (write), and Subscription
(real-time events via WebSocket). Field resolution is delegated to resolver functions,
which decouple the API schema from the underlying data sources and enable a single
endpoint to aggregate data from multiple backends --- a pattern known as
backend-for-frontend (BFF).

\textbf{Strengths.}
GraphQL eliminates both over-fetching and under-fetching simultaneously: the client
declares exactly what it needs and the server returns only that. The typed schema
functions as living documentation with interactive explorers (GraphiQL, Apollo Studio),
compile-time query validation, and type-safe client code generation. A single query can
retrieve data that would require several sequential REST requests, reducing network round
trips. Schema introspection and strong typing significantly improve developer tooling and
reduce integration errors.

\textbf{Weaknesses.}
The N+1 query problem --- where resolving N parent objects triggers N separate resolver
calls for their children --- requires the DataLoader batching pattern to avoid database
overload. HTTP caching is incompatible with the standard POST-to-single-endpoint model;
persisted queries are needed as a workaround. Deeply nested queries create
denial-of-service risk; query depth limits and complexity analysis are necessary
safeguards. Schema evolution at scale demands careful deprecation management rather than
simple version bumps.

\textbf{Stack, security, environments, and use cases.}
Server implementations include Apollo Server (Node.js), Strawberry and Graphene (Python),
graphql-java (Java), and gqlgen (Go). Security uses TLS and OAuth2/JWT; introspection
must be disabled in production; field-level authorisation is implemented in resolvers.
GraphQL is best suited to BFF layers serving diverse client populations, mobile
applications with heterogeneous data requirements, and developer platforms where
schema-driven tooling is valued. It is poorly suited to simple CRUD services with uniform
clients or architectures where HTTP caching is a first-class concern.

% ------------------------------------------------------------------ %
\subsection{gRPC}
% ------------------------------------------------------------------ %

\textbf{Architecture and principles.}
gRPC is a high-performance RPC framework using Protocol Buffers (protobuf) as both
interface definition language and binary serialisation format. Services and messages are
defined in \texttt{.proto} files; the \texttt{protoc} compiler generates typed client
stubs and server base classes in the target language. The framework runs exclusively over
HTTP/2, which provides multiplexed concurrent streams, binary framing, HPACK header
compression, and flow control. Four communication patterns are natively supported:
unary (one request, one response), server-streaming, client-streaming, and bidirectional
streaming. From the application layer, invoking a remote service is syntactically
identical to a local function call.

\textbf{Strengths.}
Performance is gRPC's defining advantage. Protocol Buffer messages use field-number tags
rather than field names, producing payloads three to ten times smaller than equivalent JSON,
with faster serialisation and deserialisation. HTTP/2 multiplexing allows multiple
concurrent RPCs over a single TCP connection without head-of-line blocking. Native
streaming enables server-side data feeds, client telemetry aggregation, and full-duplex
real-time communication --- patterns that REST and GraphQL cannot replicate natively.
Strong schema enforcement via \texttt{.proto} files eliminates serialisation mismatches,
and code generation removes boilerplate across all supported languages.

\textbf{Weaknesses.}
Binary encoding makes debugging without the schema file impossible; grpcurl and server
reflection partially mitigate this. Browsers cannot use native gRPC because HTTP/2
trailers are inaccessible via browser APIs; gRPC-Web with an Envoy proxy layer is
required. Protobuf schema evolution follows strict field-number permanence rules: numbers
cannot be reused and type changes break binary compatibility, requiring disciplined schema
management. The toolchain (protoc, plugins, schema registries) adds operational overhead
compared to text-based alternatives.

\textbf{Stack, security, environments, and use cases.}
Supported languages include Go, Java, Python, C++, C\#, Node.js, and Kotlin via plugins.
Security uses TLS for transport and mTLS for mutual identity verification in zero-trust
architectures; authorisation is implemented via server-side interceptors. gRPC is the
preferred choice for internal microservice communication, polyglot systems requiring
consistent cross-language contracts, latency-sensitive workloads, ML inference serving
(e.g.\ TensorFlow Serving), and IoT telemetry pipelines. It is unsuitable for
browser-native clients without a proxy and for public APIs targeting third-party developers
unfamiliar with protobuf tooling.

% ================================================================== %
\section{Practical Examples}
% ================================================================== %

All examples use the same domain: a \texttt{UserService} that manages user records.
This allows direct comparison at the protocol level. Complete Python server and client
implementations for all four technologies are provided in the \texttt{examples/}
directory of this project.

% ------------------------------------------------------------------ %
\subsection{SOAP / WSDL}
% ------------------------------------------------------------------ %

The WSDL excerpt below shows the type definitions and the operation contract. The
\texttt{portType} element is the WSDL equivalent of an interface: it declares the
\texttt{getUser} operation and the names of its input and output messages, each mapped
to an XSD complex type.

\begin{lstlisting}[language=XML, caption={WSDL excerpt --- type definitions and operation contract}]
<types>
  <xsd:complexType name="GetUserRequest">
    <xsd:sequence>
      <xsd:element name="userId" type="xsd:int"/>
    </xsd:sequence>
  </xsd:complexType>
  <xsd:complexType name="GetUserResponse">
    <xsd:sequence>
      <xsd:element name="userId"   type="xsd:int"/>
      <xsd:element name="username" type="xsd:string"/>
      <xsd:element name="email"    type="xsd:string"/>
      <xsd:element name="role"     type="xsd:string"/>
    </xsd:sequence>
  </xsd:complexType>
</types>
<portType name="UserPortType">
  <operation name="getUser">
    <input  message="tns:GetUserRequest"/>
    <output message="tns:GetUserResponse"/>
  </operation>
</portType>
\end{lstlisting}

The SOAP envelopes below illustrate the XML framing that every message carries,
regardless of payload size. The envelope wraps the actual business data (the user ID in
the request, the four user fields in the response) in multiple layers of XML markup,
which is the root cause of SOAP's bandwidth and parsing overhead.

\begin{lstlisting}[language=XML, caption={SOAP request and response envelopes for getUser}]
<!-- REQUEST -->
<soapenv:Envelope xmlns:soapenv="http://schemas.xmlsoap.org/soap/envelope/"
                  xmlns:usr="http://example.com/userservice">
  <soapenv:Header/>
  <soapenv:Body>
    <usr:GetUserRequest><usr:userId>42</usr:userId></usr:GetUserRequest>
  </soapenv:Body>
</soapenv:Envelope>

<!-- RESPONSE -->
<soapenv:Envelope xmlns:soapenv="http://schemas.xmlsoap.org/soap/envelope/"
                  xmlns:usr="http://example.com/userservice">
  <soapenv:Body>
    <usr:GetUserResponse>
      <usr:userId>42</usr:userId>
      <usr:username>jdupont</usr:username>
      <usr:email>j.dupont@example.com</usr:email>
      <usr:role>editor</usr:role>
    </usr:GetUserResponse>
  </soapenv:Body>
</soapenv:Envelope>
\end{lstlisting}

% ------------------------------------------------------------------ %
\subsection{REST}
% ------------------------------------------------------------------ %

The two listings below show a resource retrieval (GET) and a resource creation (POST)
with their respective responses. HTTP status codes and the \texttt{Location} header in
the POST response carry semantic meaning without requiring any parsing of the body.

\begin{lstlisting}[caption={GET /api/users/42 --- request and response}]
GET /api/users/42 HTTP/1.1
Host: api.example.com
Authorization: Bearer eyJhbGciOiJSUzI1NiJ9...

HTTP/1.1 200 OK
Content-Type: application/json

{"userId":42,"username":"jdupont","email":"j.dupont@example.com","role":"editor"}
\end{lstlisting}

\begin{lstlisting}[caption={POST /api/users --- resource creation}]
POST /api/users HTTP/1.1
Content-Type: application/json

{"username":"mmartin","email":"m.martin@example.com","role":"viewer"}

HTTP/1.1 201 Created
Location: /api/users/43
Content-Type: application/json

{"userId":43,"username":"mmartin","email":"m.martin@example.com","role":"viewer"}
\end{lstlisting}

% ------------------------------------------------------------------ %
\subsection{GraphQL}
% ------------------------------------------------------------------ %

The SDL schema defines the complete type system. The query below requests only three
fields (\texttt{username}, \texttt{email}, and the nested post titles), leaving
\texttt{userId}, \texttt{role}, and \texttt{createdAt} unselected. The response mirrors
the query structure exactly: unrequested fields are absent, and the user's posts are
returned in the same round trip --- eliminating both over-fetching and under-fetching
simultaneously.

\begin{lstlisting}[language=graphql, caption={GraphQL schema (SDL)}]
type User {
  userId: ID!   username: String!   email: String!
  role: String!   createdAt: String!   posts: [Post!]!
}
type Post   { postId: ID!   title: String!   publishedAt: String! }
type Query  { user(userId: ID!): User   users: [User!]! }
type Mutation { createUser(username: String!, email: String!, role: String!): User! }
type Subscription { userCreated: User! }
\end{lstlisting}

\begin{lstlisting}[caption={GraphQL query and server response}]
// Query (sent as HTTP POST body)
query GetUserWithPosts {
  user(userId: "42") {
    username
    email
    posts { title  publishedAt }
  }
}

// Response
{
  "data": {
    "user": {
      "username": "jdupont",
      "email": "j.dupont@example.com",
      "posts": [
        {"title": "Comparing API Paradigms", "publishedAt": "2026-02-01T14:30:00Z"}
      ]
    }
  }
}
\end{lstlisting}

% ------------------------------------------------------------------ %
\subsection{gRPC}
% ------------------------------------------------------------------ %

The \texttt{.proto} file is the authoritative contract for the service. The Python
server listing implements the three RPCs: two unary calls and one server-streaming call.
The client listing demonstrates invoking the streaming \texttt{ListUsers} RPC, which
yields user records one by one over an open HTTP/2 stream.

\begin{lstlisting}[language=protobuf, caption={user\_service.proto --- service contract}]
syntax = "proto3";
package userservice;

service UserService {
  rpc GetUser    (GetUserRequest)    returns (UserResponse);
  rpc ListUsers  (ListUsersRequest)  returns (stream UserResponse);
  rpc CreateUser (CreateUserRequest) returns (UserResponse);
}

message GetUserRequest    { int32  user_id  = 1; }
message ListUsersRequest  {}
message CreateUserRequest { string username = 1; string email = 2; string role = 3; }
message UserResponse      { int32  user_id  = 1; string username = 2;
                            string email    = 3; string role     = 4; }
\end{lstlisting}

\begin{lstlisting}[language=Python, style=codestyle,
                   caption={gRPC server --- servicer implementation (Python)}]
class UserServiceServicer(user_service_pb2_grpc.UserServiceServicer):

    def GetUser(self, request, context):
        user = USERS.get(request.user_id)
        if user is None:
            context.abort(grpc.StatusCode.NOT_FOUND, "User not found")
        return user_service_pb2.UserResponse(user_id=request.user_id, **user)

    def ListUsers(self, request, context):       # server-streaming RPC
        for uid, user in USERS.items():
            yield user_service_pb2.UserResponse(user_id=uid, **user)

    def CreateUser(self, request, context):
        nid = max(USERS) + 1
        USERS[nid] = dict(username=request.username,
                          email=request.email, role=request.role)
        return user_service_pb2.UserResponse(user_id=nid, **USERS[nid])
\end{lstlisting}

\begin{lstlisting}[language=Python, style=codestyle,
                   caption={gRPC client --- unary and streaming calls (Python)}]
with grpc.insecure_channel("localhost:50051") as channel:
    stub = user_service_pb2_grpc.UserServiceStub(channel)

    # Unary call: retrieve a single user
    r = stub.GetUser(user_service_pb2.GetUserRequest(user_id=42))
    print(f"GetUser -> {r.username}, {r.role}")

    # Server-streaming call: iterate over all users as they arrive
    for u in stub.ListUsers(user_service_pb2.ListUsersRequest()):
        print(f"  [{u.user_id}] {u.username} <{u.email}>")
\end{lstlisting}

% ================================================================== %
\section{Comparative Summary Table}
% ================================================================== %

\renewcommand{\arraystretch}{1.2}
\footnotesize
\begin{longtable}{|p{2.6cm}|p{2.85cm}|p{2.85cm}|p{2.85cm}|p{2.85cm}|}
\hline
\textbf{Dimension}
  & \textbf{SOAP / WSDL}
  & \textbf{REST}
  & \textbf{GraphQL}
  & \textbf{gRPC} \\
\hline
\endfirsthead
\hline
\textbf{Dimension}
  & \textbf{SOAP / WSDL}
  & \textbf{REST}
  & \textbf{GraphQL}
  & \textbf{gRPC} \\
\hline
\endhead
\hline
\endfoot

\textbf{Protocol / transport}
  & SOAP over HTTP, SMTP, JMS
  & HTTP/1.1 or HTTP/2
  & HTTP (POST / GET)
  & HTTP/2 only \\
\hline

\textbf{Message format}
  & XML (mandatory)
  & JSON (typical)
  & JSON
  & Protocol Buffers (binary) \\
\hline

\textbf{Schema / contract}
  & WSDL (mandatory)
  & OpenAPI (optional)
  & SDL (mandatory)
  & .proto (mandatory) \\
\hline

\textbf{Performance}
  & Low --- verbose XML
  & Moderate
  & Moderate
  & High --- binary + HTTP/2 \\
\hline

\textbf{Payload size}
  & Very large
  & Medium
  & Client-controlled
  & Very small \\
\hline

\textbf{Data fetching model}
  & Fixed per operation
  & Fixed per endpoint
  & Client-defined fields
  & Fixed per RPC \\
\hline

\textbf{Streaming support}
  & Limited (MTOM)
  & Partial (SSE)
  & Subscriptions (WS)
  & Native (4 patterns) \\
\hline

\textbf{Browser support}
  & Poor
  & Excellent
  & Excellent
  & Limited (gRPC-Web) \\
\hline

\textbf{Security model}
  & WS-Security (msg-level)
  & TLS + OAuth2 / JWT
  & TLS + OAuth2 / JWT
  & TLS + mTLS \\
\hline

\textbf{Developer usability}
  & Low (steep learning curve)
  & High
  & High (with tooling)
  & Moderate \\
\hline

\textbf{Code generation}
  & Yes (from WSDL)
  & Partial (from OpenAPI)
  & Yes (from SDL)
  & Yes (from .proto) \\
\hline

\textbf{HTTP caching}
  & Not applicable
  & Native (GET responses)
  & Persisted queries
  & Application-level only \\
\hline

\textbf{Client--server coupling}
  & Tight
  & Loose
  & Loose (schema-guided)
  & Tight (schema-governed) \\
\hline

\textbf{Versioning strategy}
  & WSDL versioning
  & URL / header versioning
  & Additive schema evolution
  & Field-number stability \\
\hline

\textbf{Typical domain}
  & Enterprise, banking, HL7
  & Public APIs, SaaS, mobile
  & BFF, SPA, mobile
  & Microservices, IoT, ML \\
\hline

\textbf{Ecosystem maturity}
  & Very mature (largely legacy)
  & Very mature
  & Maturing rapidly
  & Maturing \\
\hline

\end{longtable}
\normalsize
\renewcommand{\arraystretch}{1}

% ================================================================== %
\section{Conclusion}
% ================================================================== %

The four technologies examined in this report are not competing alternatives but
complementary tools, each suited to a distinct set of requirements. Selecting the wrong
paradigm has real costs --- wasted bandwidth, poor developer experience, or security
gaps --- while selecting the right one can significantly simplify both implementation and
operations.

SOAP remains the appropriate choice when formal contractual guarantees, message-level
security, or distributed transactional semantics carry legal or compliance weight. In
banking, insurance, and healthcare integration, these properties are not optional, and
SOAP's complexity is the price of the guarantees it provides.

REST is the rational default for any API that will be consumed by diverse, externally
managed clients. Its alignment with HTTP semantics, universal tooling support, and native
caching behaviour make it the most accessible and operationally simple option. Its
limitations around schema enforcement and data fetching efficiency are real but manageable
for the majority of public API use cases.

GraphQL addresses REST's data fetching limitations in a principled way. Its ability to
serve multiple client types from a single schema and to aggregate multiple backends into
one query layer makes it the natural choice for frontend-facing data layers in complex
applications. The server-side complexity it introduces is a reasonable trade-off when
clients are diverse and network efficiency matters.

gRPC represents the highest-performance option and is the appropriate choice for internal
service-to-service communication in polyglot microservice architectures. Its binary
encoding, HTTP/2 transport, native streaming, and cross-language code generation
provide advantages that text-based protocols cannot match for high-frequency, low-latency
communication. The accessibility trade-offs --- binary opacity, limited browser support,
protobuf toolchain overhead --- are acceptable when teams control both ends of the
connection.

In practice, mature distributed systems commonly employ all four paradigms simultaneously:
REST for the public API surface, GraphQL for the frontend aggregation layer, gRPC for
internal service communication, and SOAP where legacy integration requirements demand it.
Understanding the reasoning behind each technology, rather than treating any single one
as a universal solution, is the foundation of sound distributed systems engineering.

% ================================================================== %
\begin{thebibliography}{9}
% ================================================================== %
\bibitem{fielding2000}
Fielding, R.T. (2000).
\textit{Architectural Styles and the Design of Network-based Software Architectures}.
Doctoral dissertation, University of California, Irvine.

\bibitem{soap12}
Gudgin, M. et al.\ (2007).
\textit{SOAP Version 1.2, Part 1: Messaging Framework (Second Edition)}.
W3C Recommendation.

\bibitem{wsdl20}
Chinnici, R. et al.\ (2007).
\textit{Web Services Description Language (WSDL) Version 2.0 Part 1: Core Language}.
W3C Recommendation.

\bibitem{graphqlspec}
The GraphQL Foundation (2021).
\textit{GraphQL Specification, October 2021 Edition}.
\url{https://spec.graphql.org/October2021/}

\bibitem{grpcdocs}
Google LLC (2024).
\textit{gRPC Documentation}. \url{https://grpc.io/docs/}

\bibitem{protobufdocs}
Google LLC (2024).
\textit{Protocol Buffers Developer Guide}. \url{https://protobuf.dev/}

\bibitem{masse2011}
Masse, M. (2011).
\textit{REST API Design Rulebook}. O'Reilly Media.

\bibitem{richardson2013}
Richardson, L. and Amundsen, M. (2013).
\textit{RESTful Web APIs}. O'Reilly Media.

\end{thebibliography}

\end{document}
